Computing in the Field of Rationals

### 2012-06-21 rational_ext.tex
_

\begin{lemma}[Symmetric Monoidal (+, 0)]
  {{Pi}} is a symmetric monoidal category with tensor {{+}} and
  monoidal unit {{0}}. For any object {{b1+b2}} the required symmetry
  morphism (called in categorical literature) is
  {{swap+:b1+b2<->b2+b1}}.
\end{lemma}

>>>>

\begin{lemma}[Dagger]
  {{Pi}} is a dagger category, where every morphism {{f : b1<->b2}}
  has the adjoint {{f^{dagger}:b2 <->b1}}. The following properties
  hold (where {{g : b2 <-> b3}}): 
  \begin{enumerate}
  \item {{id^{dagger} = id : b <-> b}}
  \item {{(f (;) g)^{dagger} = f^{dagger} (;) g^{dagger}: b1 <-> b2}}
  \item {{f^{dagger dagger} = f : b1 <-> b2}}
  \end{enumerate}
\end{lemma}

\begin{lemma}[Symmetric Monoidal (+, 0)]
  {{Pi}} is a symmetric monoidal category with tensor {{+}} and
  monoidal unit {{0}}. The monoidal operation on morphisms is the
  additive composition of combinators {{c1 (+) c2}}.

  \begin{proof}
    To establish that category is monoidal one must show isomorphisms
    \begin{enumerate}
    \item {{b1 + (b2 + b3) <-> (b1 +b2) + b3}} given by {{assoc+}}
    \item {{0 + b <-> b}} given by {{zeroe}}
    \item {{b + 0 <-> b}} given by {{swap+ (;) zeroe}}
    \item We need to check that {{+}} is a bifunctor. 
      \begin{enumerate}
      \item {{id_{b1}(+)id_{b2} = id_{b1+b2} }}
      \item {{(f(+)g)(;)(j(+)k) = (f (;) j) (+) (g (;) k)}}
      \end{enumerate}
    \item We need to check the naturality of {{assoc+}}, {{zeroe}} and
      {{swap+(;)zeroe}}.
      \begin{enumerate}
      \item {{assoc+ (;) ((f (+)g) (+) h) = (f (+)(g (+) h)) (;) assoc+}}
      \item {{zeroe (;) f = (id (+) f) (;) zeroe}}
      \item {{(swap+ (;) zeroe) (;) f = (f (+) id) (;) (swap+ (;) zeroe)}}
      \end{enumerate}
    \item Satisfy certain coherence conditions which are usually
      called the pentagon and traingle axioms.
    \end{enumerate}

    The last three points require checking equality of combinators by
    writing out their derivation trees as we did in the case of
    associativity of sequential composition. To show symmetry, we need
    a braiding operation {{b1+b2 <-> b2+b1}} which is given by
    {{swap+}}.

    \begin{enumerate}
    \item The braiding must satisfy two ``hexagon axioms''. 
    \item The braiding is self inverse {{swap+_{b1+b2} = (swap+_{b2+b1})^{dagger} }}.
    \end{enumerate}

  \end{proof}

\end{lemma}

\begin{lemma}[Symmetric Monoidal (*, 1)]
  {{Pi}} is a symmetric monoidal category over the tensor {{*}} and
  unit {{1}}. The details here follow those of the {{(0, +)}} monoid
  and are skipped.
\end{lemma}

Some technical and pedantic comments are due at this point.

\begin{itemize}
\item We have established the categorical structure of {{Pi}} as a
  dagger symmetric monoidal category with two monoidal structures,
  {{(0, +)}} and {{(*, 1)}}. In Sec. \ref{sec:int} we will see that a
  {{trace}} operator can be admitted in this category without any
  change of expressiveness.

\item By definition, in {{Pi}} every morphism is an isomorphism --
  this makes {{Pi}} a groupoid.

\item The category {{Pi}} has no initial and terminal objects. The
  objects {{0}} and {{1}} would be initial and terminal if we admitted
  all functions (is in the category {{CatSet}}).

\item The category {{Pi}} has neither categorical products, nor
  categorical co-products -- {{*}} and {{+}} are merely monoidal
  tensors. This follows from the fact that injection and projection
  maps are not isomorphisms.
\end{itemize}

While we don't do so in this work, if we extend {{Pi}} with products
and co-products (say through the addition of information effects) it
is concievable that part of its structure collapses. The sort of
collapse, while catastrophic for algebraic structures (effectively
trivializing them) still retains some interest in computing, because in
computing we are interested in the specific operational nature (the
computation content, so to speak) of the morphisms. Anectodal evidence
follows from the fact several real-world programming language have
inconsistent type systems. blah blah blah...
